Automated assessment of residual plots with computer vision models

### 2024-02-05 paper/paper.tex
update the paper

\title{Automated reading of residual plots with computer vision models}

@@

However, as pointed out by \citet{li2023plot}, a primary limitation of
the lineup protocol lies in its reliance on human judgments. Unlike
conventional statistical tests that can be performed numerically and
automatically in statistical software, the lineup protocol requires
human evaluation of images. This characteristic makes it less suitable
for large-scale applications, given the associated high labor costs and
time requirements.

There is a substantial need to develop an approach that alleviates
people's workload by automating repetitive tasks and providing
standardized results in a controlled environment. The large-scale
evaluation of lineups is impractical without the use of technology and
machines.

@@

In addressing the residual reading problem, there exist multiple
configurations of the computer vision model formula. These
configurations can be categorized by two key components of the model
formula: the input and the output format.

@@

In computer vision models, the quality and relevance of the input data
greatly influence the model's capacity to generate insightful and
meaningful results. The input of the model can be designed in different
formats.

A straightforward approach involves feeding the model a vector of
residuals along with a vector of fitted values, essentially providing
all the necessary information for creating a residuals vs fitted values
plot. However, a drawback of this method is the dynamic input size, as
different fitted regression models may have varying numbers of
observations. For modern computer vision models implemented with
mainstream software like TensorFlow \citep{abadi2016tensorflow}, the
input shape is typically fixed. One solution is to pad the input vectors
with leading or trailing zeros if the input tensor expects longer
vectors, but it may fail if the input vector surpasses the designed
length. Another strategy is to summarize the residuals and fitted values
separately using histograms and utilize the counts as the input. By
controlling the number of bins in the histograms, it becomes possible to
provide fixed-length input vectors.

@@

Other possible input formats include, but are not limited to, a pair of
residual plots, a triplet, and a lineup. \citet{chopra2005learning} has
shown that computer vision models designed for image comparison can
assess whether a pair of images are similar or dissimilar. Applied to
our specific problem, we can define null plots of a fitted model to be
similar to each other, while considering actual residual plots to be
distinct from null plots of any fitted model. A triplet constitutes a
set of three images, denoted as \(image_1\), \(image_2\) and
\(image_3\). It is often used to predict whether \(image_2\) or
\(image_3\) is more similar to \(image_1\), proving particularly useful
for establishing rankings between samples. However, it's important to
note that these two approaches usually require additional considerations
regarding the loss function and, at times, non-standard training
processes due to shared weights between different convolutional blocks.

@@

In a visual test, the observer will be asked to choose one or more plots
that stand out as most distinct from others in a given lineup. To
develop a computer vision model for evaluating residual plots within the
visual inference framework, it is important to precisely define a
numerical measure of ``difference'' or ``distance'' between plots. This
distance can take the form of a basic statistical operation on pixels,
such as the sum of square differences. Alternatively, it could involve
established image similarity metrics like the Structural Similarity
Index Measure \citep{wang2004image}. The challenge lies in the fact that
metrics tailored for image comparison may not be suitable for evaluating
data plots, where only essential plot elements require assessment
\citep{chowdhury2018measuring}. Furthermore, scagnostics mentioned in
Section \ref{introduction} could be used to construct distance metrics
for data plots comparison, but the functional form still needs to be
carefully refined to accurately reflect the extent of the violations.

@@

The distance metrics proposed in this paper takes into account the fact
that we try to measure how different a residual plot is from a good
residual plot, or in other words, how different a given fitted model is
from a correctly specified model. For the classical normal linear
regression model, residuals are derived from the fitted values
\(\hat{\boldsymbol{y}}\) and observed values \(\boldsymbol{y}\). Suppose
the data generating process is known and the model is correctly
specified, by the Frisch-Waugh-Lowell theorem \citep{frisch1933partial},
residuals \(\boldsymbol{e}\) can also be written as a linear
transformation of the error \(\boldsymbol{\varepsilon}\) formulated as
\(\boldsymbol{e} = \boldsymbol{R}\boldsymbol{\varepsilon}\), where
\(\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the residual operator, \(\boldsymbol{X}\) is the design matrix,
\(\boldsymbol{I}_n\) is a \(n\) by \(n\) identity matrix, and \(n\) is
the number of observations.

@@

When fitting a linear regression model, the solver will force
\(\sum_{i=1}^{n} e_i = 0\), making any residual value to be a linear
combination of the remaining \(n - 1\) residuals. This effectively means
\(rank(\boldsymbol{R}) = n - 1 < n\) and \(Q\) becomes a degenerate
multivariate distribution. To capture the characteristics of \(Q\), such
as moments, we can simulate a large numbers of
\(\boldsymbol{\varepsilon}\) and transform it to \(\boldsymbol{e}\) to
get the empirical estimates. For simplicity, we replaced
\(\boldsymbol{R}\) with a full-rank diagonal matrix
\(diag(\boldsymbol{R})\), where \(diag(.)\) set the non-diagonal entries
of a matrix to zeros. The resulting distribution for \(\boldsymbol{Q}\)
is \(N(\boldsymbol{0}_n, diag(\boldsymbol{R}\sigma^2))\).

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different to \(Q\). For example, if
the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but not included in \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different to \(Q\)
because the residual operator obtained from the fitted model will not be
the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate lognormal distribution, the empirical residual distribution
will usually be skewed and has a long tail.

@@

This distance metric was first proposed in \citet{li2023plot}. It was
mainly designed for measuring the effect size of non-linearity and
heteroskedasticity in a residual plot. \citet{li2023plot} showed that,
for a classical normal linear regression model that omits a necessary
higher-order predictors \(\boldsymbol{Z}\), and incorrectly assumes
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\)
while in fact
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})\),
\(Q\) can be represented as
\(N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, diag(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}'))\).
Note that the variance-covariance matrix is replaced with the diagonal
matrix to ensure it is a full-rank matrix.

@@

\begin{align}
\label{eq:kl-2}
D_{KL} &= \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{W})|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\text{diag}(\boldsymbol{W})^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z'(diag(\boldsymbol{W}))^{-1}\boldsymbol{\mu}_z\right),
\end{align}

\noindent where
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\),
and \(\boldsymbol{W} = \boldsymbol{R}\boldsymbol{V}\boldsymbol{R}'\).
The assumed error variance \(\sigma^2\) is set to be
\(tr(\boldsymbol{V})/n\), which is the expectation of the estimated
variance.

@@

In order to evaluate equation \ref{eq:kl-1} in a practically computable
manner, the elements of \(\boldsymbol{e}\) are assumed to be independent
of each other. This assumption solves both of the issues mentioned
above. First, we no longer need to integrate over \(n\) random
variables. The result of equation \ref{eq:kl-1} is now the sum of the
Kullback-Leibler divergence evaluated for each individual residual
thanks for the independence assumption. Second, it is not required to
know the joint probability density \(p(\boldsymbol{e})\) any more.
Instead, the evaluation of Kullback-Leibler divergence for an individual
residual relies on the knowledge of the marginal density \(p_i(e_i)\),
where \(e_i\) is the \(i\)-th residual for \(i = 1, ..., n\). This is
much easier to estimate through simulation.

The algorithm for computing equation \ref{eq:kl-1} starts from
simulating \(m\) sets of \(\boldsymbol{\varepsilon}\) according to the
error distribution. The simulated errors are stored in a matrix
\(\boldsymbol{A}\) with \(n\) rows and \(m\) columns. So each column of
\(\boldsymbol{A}\) is a set of realization values of
\(\boldsymbol{\varepsilon}\). Then, we can get \(m\) sets of
\(\boldsymbol{e}\) stored in the matrix \(\boldsymbol{B}\) by applying
the residual operator \(\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}\).
Furthermore, kernel density estimation (KDE) with Gaussian kernel and
optimal bandwidth selected by the Silverman's rule of thumb
\citep{silverman2018density} is applied on each row of \(B\) to estimate
\(p_i(e_i)\) for \(i = 1, ..., n\). The KDE computation is done by the
\texttt{density} function in R.

Since the Kullback-Leibler divergence can be viewed as the expectation
of the log-likelihood ratio between distribution \(P\) and distribution
\(Q\) evaluated on distribution \(P\), we can reuse the simulated
residuals in matrix \(\boldsymbol{B}\) to estimate the expectation by
the sample mean. With the independence assumption, for non-normal \(P\),
\(D_{KL}\) can be estimated by

\begin{align}
\label{eq:kl-3}
\widehat{D_{KL}} &= \sum_{i = 1}^{n} \widehat{D_{KL}^{(i)}}, \\
\widehat{D_{KL}^{(i)}} &= \frac{1}{m}\sum_{j = 1}^{m} log\frac{\hat{p_i}(B_{ij})}{q(B_{ij})},
\end{align}

\noindent where \(\widehat{D_{KL}^{(i)}}\) is the estimator of the
Kullback-Leibler divergence for an individual residual \(e_i\),
\(B_{ij}\) is the \(i\)-th row and \(j\)-th column entry of the matrix
\(B\), \(\hat{p_i}(.)\) is the kernel density estimator of \(p_i(.)\),
\(q(.)\) is the normal density function with mean zero and an assumed
variance estimated as
\(\widehat{\sigma^2} = \sum_{b \in vec(B)}(b - \sum_{b \in vec(B)} b/nm)^2/(nm - 1)\),
and \(vec(.)\) is the vectorization operator which turns a
\(n \times m\) matrix into a \(nm \times 1\) column vector by stacking
the columns of the matrix on top of each other.

@@

\begin{equation}
\label{eq:d-approx}
\hat{D} = f(V(\hat{\boldsymbol{e}}, \hat{\boldsymbol{y}})),
\end{equation}

\noindent where \(\hat{\boldsymbol{e}}\) is a vector of residuals
obtained from the fitted model rather than a vector of random variables,
\(\hat{\boldsymbol{y}}\) is the fitted values also obtained from the
fitted model, \(V(.)\) is a plotting function that generates a residuals
vs fitted values plot with fixed aesthetic and then saves it as an image
with \(h \times w\) pixels and three colour channels, and \(f(.)\) is a
computer vision model which takes an \(h \times w\) image as input and
predicts the distance in the domain \([0, +\infty)\).

@@

Theoretically, the distance \(D\) for a correctly specified model is
\(0\), because \(P\) will be the same as \(Q\). However, the computer
vision model may not necessary predict \(0\) for a null plot. Using
Figure \ref{fig:false-finding} as an example, it contains a visual
pattern which is an indication of heteroskedasticity. We would not
expect the model to be able to magically tell if the suspicious pattern
is caused by the skewed distribution of the fitted values or the
existence of heteroskedasticity. Assume we only train the model with
this single image, while \(50\)\% of the time the label is \(0\), and
\(50\)\% of the time the label is some constant \(c > 0\). Then, a
perfect model will predict \(0 < \hat{D} < c\) to achieve a minimum
loss. Some null plots could have outliers or strong visual patterns due
to randomness, and a reasonable model will try to summarise those
information into the prediction, resulting in \(\hat{D} > 0\).

This property is not an issue if \(\hat{D} \gg 0\) for which the visual
signal of the residual plot is very strong, and we usually do not need
any further examination of the significance of the result. However, if
the visual pattern is weak or moderate, having \(\hat{D}\) will not be
sufficient to determine if the null hypothesis that the model is
correctly specified should be rejected.

To solve this issue while aligning with the principle of visual
inference, the approximated distance \(\hat{D}\) can be defined as a
test statistic. And the null distribution of this statistic can be
approximated by the empirical distribution of \(\widehat{D_{null}}\) for
a large amount of null plots of a fitted model. The procedure involves
applying the residual rotation technique \citep{buja2009statistical} on
the fitted model to obtain null residuals. The null residuals are then
used to make null plots and fed into the model. The predicted distance
are used to construct an empirical distribution.

@@

Let \(\widehat{D_{null}^{(i)}}\) be the predicted distance of the
\(i\)-th null plots, where \(i = 1,...,m\) and \(m \in \mathbb{N}^+\) is
a sufficiently large number. Quantiles of the null distribution can then
be estimated using the sample quantiles available in statistical
software such as R. The details of the sample quantile computation can
be found in \citet{hyndman1996sample}. In statistical testing, analysts
often care about certain quantiles of the null distribution, such as
90\% quantile, 95\% quantile and 99\% quantile. These quantiles are used
as thresholds to decide if \(H_0\) needs to be rejected. For example, if
\(\hat{D}\) is greater than and equal to the 95\% sample quantile of the
null distribution, we could say the predicted distance for the actual
residual plot is significantly different from the predicted distance for
null plots with 95\% significance level. Based on our experience, in
order to get a stable estimate of the 95\% quantile, \(m\) usually needs
to be at least \(100\). And if the null distribution has a long tail,
more null plots will be needed. Alternatively, a p-value can be used to
represents the probability of observing an event equally or more extreme
than the given event under \(H_0\), and it can be estimated by
\(1/m\sum_{i=1}^{m}I\left(\widehat{D_{null}^{(i)}} \geq \hat{D}\right)\).

@@

\begin{itemize}
\tightlist
\item
  What is the assumption of bootstrapping
\item
  Any drawback: loss of information
\item
  For each bootstrapped fitted model, we can get a bootstrapped vss and
  construct a new 95\% quantile for the null distribution, and check if
  H\_0 should be rejected
\item
  However, it is very expensive to construct a new 95\% quantile for
  each bootstrapped fitted model
\item
  The 95\% quantile constructed using the original fitted model can be
  considered as an estimate of those new 95\% quantiles
\item
  We borrow information from the original fitted model
\item
  We can check how many bootstrapped fitted model should be rejected
  (like the percentage)
\item
  This gives an overall estimate of how often the assumed regression
  model are considered to be incorrect if the data can be obtained
  repetitively from the same data generating process
\item
  If this percentage is relatively large, it gives the analyst an
  warning that the regression model is inappropriate to the data
\end{itemize}

@@

\begin{align}
\label{eq:data-sim}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
\boldsymbol{z} &= He_j(g(\boldsymbol{x}_1, 2)), \\
\boldsymbol{w} &= He_j(g(\boldsymbol{x}_2, 2)), \\
\boldsymbol{k} &= \sqrt{\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^2},
\end{align}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}_1\),
\(\boldsymbol{x}_2\), \(\boldsymbol{z}\), \(\boldsymbol{w}\),
\(\boldsymbol{k}\) and \(\boldsymbol{\varepsilon}\) are vectors of size
\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
\(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\) are two independent
predictors, \(He_j(.)\) is the \(j\)th-order probabilist's Hermite
polynomials \citep{hermite1864nouveau}, the \(\sqrt{(.)}\) and \((.)^2\)
operators are element-wise operators, \(\odot\) is the Hadamard product,
and \(g(., k)\) is a scaling function to enforce the support of the
random vector to be \([-k, k]^n\) defined as

@@

The model is trained on a high-performance computing platform owned by
Monash university with TensorFlow (ref here) and Keras (ref here).
During the training process, 80\% of the training data are used as the
actual training data, and 20\% of the training data are reserved as
validation data. We let the Bayesian optimization tuner runs 100 trials
to find the best hyperparameter values according to the validation root
mean square error. The tuner will restore the best epoch of the best
model from the trials. Early stopping is also applied such that the
training process will halt if the validation root mean square error does
not improve for 50 epochs. The maximum allowed epochs is 2000 where no
models reach.

>>>>

\title{Automated assessment of residual plots with computer vision
models}

@@

However, as pointed out by \citet{li2023plot}, a primary limitation of
the lineup protocol lies in its reliance on human judgments. Unlike
conventional statistical tests that can be performed computationally in
statistical software, the lineup protocol requires human evaluation of
images. This characteristic makes it less suitable for large-scale
applications, given the associated high labor costs and time
requirements.

There is a substantial need to develop an approach that alleviates
analysts' workload by automating repetitive tasks and providing
standardized results in a controlled environment. The large-scale
evaluation of lineups is impractical without the use of technology and
machines.

@@

There are various configurations of the computer vision model that can
be used to assess residual plots. We discuss these configurations below
based on two key components of the model formula: the input and the
output format.

@@

Deep learning models are in general very sensitive to the input data.
The quality and relevance of the input data greatly influence the
model's capacity to generate insightful and meaningful results.

A straightforward approach involves feeding a vector of residuals along
with a vector of fitted values, essentially providing all the necessary
information for creating a residuals vs fitted values plot. However, a
drawback of this method is the dynamic input size, which changes based
on the number of observations. For modern computer vision models
implemented in mainstream software like TensorFlow
\citep{abadi2016tensorflow}, the input shape is typically fixed. One
solution is to pad the input vectors with leading or trailing zeros when
the input tensor expects longer vectors, but it may fail if the input
vector surpasses the designed length. Another strategy is to summarize
the residuals and fitted values separately using histograms and utilize
the counts as the input. By controlling the number of bins in the
histograms, it becomes possible to provide fixed-length input vectors.
Still, since the histogram only captures the marginal distribution of
residuals and fitted values respectively, it can not be used to
differentiate visual patterns with same marginal distributions but
different joint distributions.

@@

Providing multiple residual plots to the model, such as a pair of plots,
a triplet or a lineup is also a possible option.
\citet{chopra2005learning} has shown that computer vision models
designed for image comparison can assess whether a pair of images are
similar or dissimilar. Applied to our specific problem, we can define
null plots of a fitted model to be similar to each other, while
considering actual residual plots to be distinct from null plots of any
fitted model. A triplet constitutes a set of three images, denoted as
\(image_1\), \(image_2\) and \(image_3\). It is often used to predict
whether \(image_2\) or \(image_3\) is more similar to \(image_1\),
proving particularly useful for establishing rankings between samples.
For this setup, we can apply the same criteria to define similarity
between images. However, it's important to note that these two
approaches usually require additional considerations regarding the loss
function and, at times, non-standard training processes due to shared
weights between different convolutional blocks.

@@

In a visual test, the observer will be asked to choose one or more plots
that stand out as most distinct from others in a given lineup. To
develop a computer vision model for assessing residual plots within the
visual inference framework, it is important to precisely define a
numerical measure of ``difference'' or ``distance'' between plots. This
distance can take the form of a basic statistical operation on pixels,
such as the sum of square differences. Alternatively, it could involve
established image similarity metrics like the Structural Similarity
Index Measure \citep{wang2004image}. The challenge lies in the fact that
metrics tailored for image comparison may not be suitable for evaluating
data plots, where only essential plot elements require assessment
\citep{chowdhury2018measuring}. Furthermore, scagnostics mentioned in
Section \ref{introduction} could be used to construct distance metrics
for data plots comparison, but the functional form still needs to be
carefully refined to accurately reflect the extent of the violations.

@@

The distance metrics proposed in this paper takes into account the fact
that we try to measure how different a residual plot is from a good
residual plot, or in other words, how different a given fitted model is
from a correctly specified model. For the classical normal linear
regression model, residuals \(\boldsymbol{e}\) are derived from the
fitted values \(\hat{\boldsymbol{y}}\) and observed values
\(\boldsymbol{y}\). Suppose the data generating process is known and the
regression model is correctly specified, by the Frisch-Waugh-Lowell
theorem \citep{frisch1933partial}, residuals \(\boldsymbol{e}\) can also
be treated as random variables and written as a linear transformation of
the error \(\boldsymbol{\varepsilon}\) formulated as
\(\boldsymbol{e} = \boldsymbol{R}\boldsymbol{\varepsilon}\), where
\(\boldsymbol{R}=\boldsymbol{I}_n -\boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'\)
is the residual operator and an idempotent matrix, \(\boldsymbol{X}\) is
the design matrix, \(\boldsymbol{I}_n\) is a \(n\) by \(n\) identity
matrix, and \(n\) is the number of observations.

@@

In ordinary least square, the minimization of the sum of square
residuals implies \(\sum_{i=1}^{n} e_i = 0\), making any residual value
to be a linear combination of the remaining \(n - 1\) residuals. This
effectively means \(\text{rank}(\boldsymbol{R}) = n - 1 < n\) and \(Q\)
is a degenerate multivariate distribution. To capture the
characteristics of \(Q\), such as moments, we can simulate a large
numbers of \(\boldsymbol{\varepsilon}\) and transform it to
\(\boldsymbol{e}\) to get the empirical estimates. For simplicity, we
replaced the variance-covariance matrix of residuals
\(\text{cov}(\boldsymbol{e}, \boldsymbol{e}) = \boldsymbol{R}\sigma^2\boldsymbol{R}' = \boldsymbol{R}\sigma^2\)
with a full-rank diagonal matrix
\(\text{diag}(\boldsymbol{R}\sigma^2)\), where \(\text{diag}(.)\) sets
the non-diagonal entries of a matrix to zeros. The resulting
distribution for \(\boldsymbol{Q}\) is
\(N(\boldsymbol{0}_n, \text{diag}(\boldsymbol{R}\sigma^2))\).

Distribution \(Q\) is derived from the correctly specified model.
However, if the model is misspecified, then the actual distribution of
residuals denoted as \(P\), will be different from \(Q\). For example,
if the data generating process contains variables correlated with any
column of \(\boldsymbol{X}\) but not included in \(\boldsymbol{X}\),
causing an omitted variable problem, \(P\) will be different from \(Q\)
because the residual operator obtained from the fitted model will not be
the same as \(\boldsymbol{R}\). Besides, if the
\(\boldsymbol{\varepsilon}\) follows a non-normal distribution such as a
multivariate lognormal distribution, \(P\) will usually be skewed and
has a long tail.

@@

This distance metric was first proposed in \citet{li2023plot}. It was
mainly designed for measuring the effect size of non-linearity and
heteroskedasticity in a residual plot. \citet{li2023plot} showed that,
for a classical normal linear regression model that omits a necessary
higher-order predictors \(\boldsymbol{Z}\), and incorrectly assumes
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n,\sigma^2\boldsymbol{I}_n)\)
while in fact
\(\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{V})\)
with \(\boldsymbol{V}\) being an arbitrary symmetric positive
semi-definite matrix, \(Q\) can be represented as
\(N(\boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z, \text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}))\).
Note that the variance-covariance matrix is replaced with the diagonal
matrix to ensure it is a full-rank matrix.

@@

\begin{align}
\label{eq:kl-2}
D_{KL} &= \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{W})|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\text{diag}(\boldsymbol{W})^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z'(\text{diag}(\boldsymbol{W}))^{-1}\boldsymbol{\mu}_z\right),
\end{align}

\noindent where
\(\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z\),
and \(\boldsymbol{W} = \boldsymbol{R}\boldsymbol{V}\boldsymbol{R}\). The
assumed error variance \(\sigma^2\) is set to be
\(\text{tr}(\boldsymbol{V})/n\), which is the expectation of the
estimated variance.

@@

In order to approximate \(D_{KL}\) in a practically computable manner,
the elements of \(\boldsymbol{e}\) are assumed to be independent of each
other. This assumption solves both of the issues mentioned above. First,
we no longer need to integrate over \(n\) random variables. The result
of equation \ref{eq:kl-1} is now the sum of the Kullback-Leibler
divergence evaluated for each individual residual thanks for the
independence assumption. Second, it is not required to know the joint
probability density \(p(\boldsymbol{e})\) any more. Instead, the
evaluation of Kullback-Leibler divergence for an individual residual
relies on the knowledge of the marginal density \(p_i(e_i)\), where
\(e_i\) is the \(i\)-th residual for \(i = 1, ..., n\). This is much
easier to approximate through simulation. It is also worth mentioning
that this independence assumption generally will not hold, since
\(\text{cov}(e_i, e_j) \neq 0\) if \(\boldsymbol{R}_{ij} \neq 0\) for
any \(1 \leq i < j \leq n\), but its existence is essential for reducing
the computational cost.

The algorithm for approximating equation \ref{eq:kl-1} starts from
simulating \(m\) sets of \(\boldsymbol{\varepsilon}\) according to the
error distribution. The simulated errors are stored in a matrix
\(\boldsymbol{A}\) with \(n\) rows and \(m\) columns, where each column
of \(\boldsymbol{A}\) is a set of realized values of
\(\boldsymbol{\varepsilon}\). Then, we can get \(m\) sets of realized
values of \(\boldsymbol{e}\) stored in the matrix \(\boldsymbol{B}\) by
applying the residual operator
\(\boldsymbol{B} = \boldsymbol{R}\boldsymbol{A}\). Furthermore, kernel
density estimation (KDE) with Gaussian kernel and optimal bandwidth
selected by the Silverman's rule of thumb \citep{silverman2018density}
is applied on each row of \(B\) to estimate \(p_i(e_i)\) for
\(i = 1, ..., n\). The KDE computation is done by the \texttt{density}
function in R.

Since the Kullback-Leibler divergence can be viewed as the expectation
of the log-likelihood ratio between distribution \(P\) and distribution
\(Q\) evaluated on distribution \(P\), we can reuse the simulated
residuals in matrix \(\boldsymbol{B}\) to estimate the expectation by
the sample mean. With the independence assumption, for non-normal \(P\),
\(D_{KL}\) can be approximated by

\begin{align}
\label{eq:kl-3}
D_{KL} &\approx \sum_{i = 1}^{n} \hat{D}_{KL}^{(i)}, \\
\hat{D}_{KL}^{(i)} &= \frac{1}{m}\sum_{j = 1}^{m} log\frac{\hat{p_i}(B_{ij})}{q(B_{ij})},
\end{align}

\noindent where \(\hat{D}_{KL}^{(i)}\) is the estimator of the
Kullback-Leibler divergence for an individual residual \(e_i\),
\(B_{ij}\) is the \(i\)-th row and \(j\)-th column entry of the matrix
\(B\), \(\hat{p_i}(.)\) is the kernel density estimator of \(p_i(.)\),
\(q(.)\) is the normal density function with mean zero and an assumed
variance estimated as
\(\widehat{\sigma^2} = \sum_{b \in vec(B)}(b - \sum_{b \in vec(B)} b/nm)^2/(nm - 1)\),
and \(vec(.)\) is the vectorization operator which turns a
\(n \times m\) matrix into a \(nm \times 1\) column vector by stacking
the columns of the matrix on top of each other.

@@

\begin{equation}
\label{eq:d-approx}
\hat{D} = f_{CV}(V_{h \times w}(\boldsymbol{e}, \hat{\boldsymbol{y}})),
\end{equation}

\noindent where \(V_{h \times w}(.)\) is a plotting function that saves
a residuals vs fitted values plot with fixed aesthetic as an image with
\(h \times w\) pixels and three colour channels, \(f_{CV}(.)\) is a
computer vision model which takes an \(h \times w\) image as input and
predicts the distance in the domain \([0, +\infty)\).

@@

Theoretically, the distance \(D\) for a correctly specified model is
\(0\), because \(P\) will be the same as \(Q\). However, the computer
vision model may not necessary predict \(0\) for a null plot. Using
Figure \ref{fig:false-finding} as an example, it contains a visual
pattern which is an indication of heteroskedasticity. We would not
expect the model to be able to magically tell if the suspicious pattern
is caused by the skewed distribution of the fitted values or the
existence of heteroskedasticity. Additionally, some null plots could
have outliers or strong visual patterns due to randomness, and a
reasonable model will try to summarise those information into the
prediction, resulting in \(\hat{D} > 0\).

This property is not an issue if \(\hat{D} \gg 0\) for which the visual
signal of the residual plot is very strong, and we usually do not need
any further examination of the significance of the result. However, if
the visual pattern is weak or moderate, having \(\hat{D}\) will not be
sufficient to determine if \(H_0\) that the model is correctly specified
should be rejected.

To solve this issue while aligning with the principle of visual
inference, the approximated distance \(\hat{D}\) can be viewed as a test
statistic. And the null distribution of this statistic can be
approximated by the empirical distribution of \(\hat{D}_{null}\) for a
large amount of null plots of a fitted model. The procedure involves
applying the residual rotation technique \citep{buja2009statistical} on
the fitted model to obtain null residuals. The null residuals are then
used to make null plots and fed into the model to get predictions. The
empirical distribution is constructed with the predicted distance.

There are two types of error associated with the approximation of the
null distribution. These include the sampling error and the estimation
error for model parameters. Increasing the number of null plots within
the distribution approximation can effectively reduce sampling error.
However, the estimation error for model parameters remains irreducible.
The residual rotation technique operates under the assumption of a
correct fitted model, using the estimated variance instead of the
unknown true variance to rescale rotated residuals. Consequently, under
\(H_0\), the true distribution of \(\hat{D}_{null}\) will slightly
deviate from the approximated null distribution.

@@

Let \(\hat{D}_{null}^{(i)}\) be the predicted distance of the \(i\)-th
null plots, where \(i = 1,...,n_{null}\) and
\(n_{null} \in \mathbb{N}^+\) is a sufficiently large number. Quantiles
of the null distribution can then be estimated using the sample
quantiles available in statistical software such as R. The details of
the sample quantile computation can be found in
\citet{hyndman1996sample}. In statistical testing, analysts often care
about certain quantiles of the null distribution, such as the 90\%
quantile, the 95\% quantile and the 99\% quantile. These quantiles are
used as thresholds to decide if \(H_0\) needs to be rejected. For
example, if \(\hat{D}\) is greater than and equal to the 95\% sample
quantile of the null distribution \(Q_{null}(0.95)\), we could say the
predicted distance for the actual residual plot is significantly
different from the predicted distance for null plots with 95\%
significance level. Based on our experience, in order to get a stable
estimate of the 95\% quantile, \(n_{null}\) usually needs to be at least
\(100\). And if the null distribution has a long tail, more null plots
will be needed. Alternatively, a p-value can be used to represents the
probability of observing an event equally or more extreme than the given
event under \(H_0\), and it can be estimated by
\(1/m\sum_{i=1}^{n_{null}}I\left(\hat{D}_{null}^{(i)} \geq \hat{D}\right)\).

@@

Bootstrap is often employed in linear regression when conducting
inference for estimated parameters. It is typically done by sampling
individual cases with replacement and refitting the regression model. If
the observed data accurately reflects the true distribution of the
population, the bootstrapped estimates can be used to measure the
variability of the parameter estimate without making strong
distributional assumptions about the data generating process.

Similarly, bootstrap can be applied on the approximated distance
\(\hat{D}\). For each refitted model \(M_{boot}^{(i)}\), there will be
an associated residual plot \(V_{boot}^{(i)}\) which can be fed into the
computer vision model to obtain \(\hat{D}_{boot}^{(i)}\), where
\(i = 1,...,n_{boot}\), and \(n_{boot}\) is the number of bootstrapped
samples. If we are interested in the variation of \(\hat{D}\), we can
use \(\hat{D}_{boot}^{(i)}\) to estimate a confidence interval.

Alternatively, since each \(M_{boot}^{(i)}\) has a set of estimated
coefficients \(\hat{\boldsymbol{\beta}}_{boot}^{(i)}\) and an estimated
variance \(\hat{\sigma^2}_{boot}^{(i)}\), a new approximated null
distribution can be construed and the corresponding 95\% sample quantile
\(Q_{boot}^{(i)}(0.95)\) can be computed. Then, if
\(\hat{D}_{boot}^{(i)} \geq Q_{boot}^{(i)}(0.95)\), \(H_0\) will be
rejected for \(M_{boot}^{(i)}\). The ratio of rejected
\(M_{boot}^{(i)}\) among all the refitted models provides an indication
of how often the assumed regression model are considered to be incorrect
if the data can be obtained repetitively from the same data generating
process. But this approach is computationally very expensive because it
requires \(n_{boot} \times n_{null}\) times of residual plot assessment.
In practice, \(Q_{null}(0.95)\) will be used to replace
\(Q_{boot}^{(i)}(0.95)\) in the computation.

@@

\begin{align}
\label{eq:data-sim}
\boldsymbol{y} &= \boldsymbol{1}_n + \boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 + \beta_2(\boldsymbol{z} + \beta_1\boldsymbol{w}) + \boldsymbol{k} \odot \boldsymbol{\varepsilon}, \\
\boldsymbol{z} &= \text{He}_j(g(\boldsymbol{x}_1, 2)), \\
\boldsymbol{w} &= \text{He}_j(g(\boldsymbol{x}_2, 2)), \\
\boldsymbol{k} &= \sqrt{\boldsymbol{1}_n + b(2 - |a|)(\boldsymbol{x}_1 + \beta_1\boldsymbol{x}_2 - a\boldsymbol{1}_n)^2},
\end{align}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}_1\),
\(\boldsymbol{x}_2\), \(\boldsymbol{z}\), \(\boldsymbol{w}\),
\(\boldsymbol{k}\) and \(\boldsymbol{\varepsilon}\) are vectors of size
\(n\), \(\boldsymbol{1}_n\) is a vector of ones of size \(n\),
\(\boldsymbol{x}_1\) and \(\boldsymbol{x}_2\) are two independent
predictors, \(\text{He}_j(.)\) is the \(j\)th-order probabilist's
Hermite polynomials \citep{hermite1864nouveau}, the \(\sqrt{(.)}\) and
\((.)^2\) operators are element-wise operators, \(\odot\) is the
Hadamard product, and \(g(., k)\) is a scaling function to enforce the
support of the random vector to be \([-k, k]^n\) defined as

@@

The model is trained on the M3 high-performance computing platform (ref
here) with TensorFlow (ref here) and Keras (ref here). During the
training process, 80\% of the training data are used as the actual
training data, and 20\% of the training data are reserved as validation
data. We let the Bayesian optimization tuner runs 100 trials to find the
best hyperparameter values according to the validation root mean square
error. The tuner will restore the best epoch of the best model from the
trials. Early stopping is also applied such that the training process
will halt if the validation root mean square error does not improve for
50 epochs. The maximum allowed epochs is 2000 where no models reach.
